RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-08-21 manuscripts/manuscript.Rmd
removed misplaced single quote after boettiger paper reference
inserted () parens in references to function calls in paragraphs in the manusript, #58

NeXML can not only contain multiple phylogenetic trees in a single document,
but can also group sets of trees into separate blocks (e.g. as resulting from
different workflows). The `get_trees` function attempts to return the most sensible 
R object for NeXML: if only a single tree is available, it will return an `ape::phylo` 
object; if multiple trees are available in a single `trees` block, it will return
an `ape::multiPhylo` object (a list of `ape::phylo` objects).  If multiple
blocks are available, it will return a list of `multiPhylo` blocks.
While most users will find this behavior convenient, there are times
(such as looping over many NeXML files) when having a variable return
type is undesirable.  To always get back a list of `multiPhylo` objects,
even when only one tree is available, use `get_trees_list()`.

@@

Previously, this was possible using the `treebase` package [@Boettiger_2012"]
for R, which relied on the NEXUS parser provided in the `ape` package.
Due to inconsistencies present in the NEXUS standard [@Vos_2012], 
a significant number <!-- State explicitly --> of TreeBASE NEXUS
files could not be read in this way.  Using the more precisely defined
NeXML standard, every valid NeXML document on TreeBASE can now successfully
be parsed to R objects.

@@

Like NEXUS files, NeXML documents can also contain character data.  While
most researchers will be familiar with the use of character matrices in
NEXUS files as a container for molecular sequence (or morphological) data 
used to infer the phylogenetic tree, we find this ideal for character data used in
comparative methods popular in R. Methods to extract character data
work much like those for the phylogenetic data. Here we read in an example
NeXML file that contains both a phylogeny and character data. 
`get_characters` returns a `data.frame` with columns as characters and rows as
taxa:

@@

The first argument to `nexml_write` can be a `ape::phylo` phylogeny,
`multiPhylo` list of phylogenies, or a `nexml` object -- anything that
can be coerced into the `nexml` class.  As we have seen, NeXML files
can contain multiple phylogenetic trees, as well as character matrices.
We can add both trees  and character data using `nexml_write`,

@@

NeXML files can also be built up bit by bit using the `add_characters`,
`add_trees`, and `add_meta`.  Each of these functions can add data to an
existing `nexml` object or create a new nexml object if none is provided.
Here we first create a `nexml` object containing the phylogeny data,
and then add the character data to it:

@@

To adjust these and other common metadata fields that should 
be part of any data deposition, `RNeXML` provides the `add_basic_meta()`
function.  This function does not assume any familiarity with the concepts
of namespaces or other more advanced metadata tools discussed later.  The
`add_basic_meta` takes an existing `nexml` object (or will create a new one
if none is provided, like the other `add_` functions), and any of the following
fields: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
`citation`.  Additional fields may be added to subsequent versions, see
`?add_basic_metadata` for the most up-to-date information.  Here we create
a new `nexml` object with the basic metadata for the `bird.orders` data
provided in the `ape` package:

@@

The `taxize_nexml` function uses the R package `taxize` 
[@Chamberlain_2013] to check each taxon label against the NCBI database.
If a unique match is found, a metadata annotation is added to the taxon
providing the NCBI identification number to the taxonomic unit.  If no
match is found, the user is warned to check for possible typographic
errors in the taxonomic labels provided.

@@

From the list of namespaces in `prefixes`, we can follow links a handful
of established vocabularies and find terms that are not included in
`add_basic_meta`.  For instance, by resolving the location identified
by the `prism` prefix,

@@

We can add this annotation to our existing `birds` NeXML file using the
`add_meta` function.  Because we do not specify a level, it is added to
the root node, referring to the NeXML file as a whole.

@@

The `get_trees` function can be used to return an `ape::phylo` tree as
usual.  `RNeXML` automatically detects the `simmap` reconstruction data
and returns includes this in a `maps` element of the `ape::phylo` object,
for use with other `phytools` functions.

>>>>

NeXML can not only contain multiple phylogenetic trees in a single document,
but can also group sets of trees into separate blocks (e.g. as resulting from
different workflows). The `get_trees()` function attempts to return the most sensible 
R object for NeXML: if only a single tree is available, it will return an `ape::phylo` 
object; if multiple trees are available in a single `trees` block, it will return
an `ape::multiPhylo` object (a list of `ape::phylo` objects).  If multiple
blocks are available, it will return a list of `multiPhylo` blocks.
While most users will find this behavior convenient, there are times
(such as looping over many NeXML files) when having a variable return
type is undesirable.  To always get back a list of `multiPhylo` objects,
even when only one tree is available, use `get_trees_list()`.

@@

Previously, this was possible using the `treebase` package [@Boettiger_2012]
for R, which relied on the NEXUS parser provided in the `ape` package.
Due to inconsistencies present in the NEXUS standard [@Vos_2012], 
a significant number <!-- State explicitly --> of TreeBASE NEXUS
files could not be read in this way.  Using the more precisely defined
NeXML standard, every valid NeXML document on TreeBASE can now successfully
be parsed to R objects.

@@

Like NEXUS files, NeXML documents can also contain character data.  While
most researchers will be familiar with the use of character matrices in
NEXUS files as a container for molecular sequence (or morphological) data 
used to infer the phylogenetic tree, we find this ideal for character data used in
comparative methods popular in R. Methods to extract character data
work much like those for the phylogenetic data. Here we read in an example
NeXML file that contains both a phylogeny and character data. 
`get_characters()` returns a `data.frame` with columns as characters and rows as
taxa:

@@

The first argument to `nexml_write()` can be a `ape::phylo` phylogeny,
`multiPhylo` list of phylogenies, or a `nexml` object -- anything that
can be coerced into the `nexml` class.  As we have seen, NeXML files
can contain multiple phylogenetic trees, as well as character matrices.
We can add both trees and character data using `nexml_write()`,

@@

NeXML files can also be built up bit by bit using `add_characters()`,
`add_trees()`, and `add_meta()`.  Each of these functions can add data to an
existing `nexml` object or create a new nexml object if none is provided.
Here we first create a `nexml` object containing the phylogeny data,
and then add the character data to it:

@@

To adjust these and other common metadata fields that should 
be part of any data deposition, `RNeXML` provides the `add_basic_meta()`
function.  This function does not assume any familiarity with the concepts
of namespaces or other more advanced metadata tools discussed later.  The
`add_basic_meta()` function takes an existing `nexml` object (or will create a new one
if none is provided, like the other `add_` functions), and any of the following
fields: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
`citation`.  Additional fields may be added to subsequent versions, see
`?add_basic_metadata` for the most up-to-date information.  Here we create
a new `nexml` object with the basic metadata for the `bird.orders` data
provided in the `ape` package:

@@

The `taxize_nexml()` function uses the R package `taxize` 
[@Chamberlain_2013] to check each taxon label against the NCBI database.
If a unique match is found, a metadata annotation is added to the taxon
providing the NCBI identification number to the taxonomic unit.  If no
match is found, the user is warned to check for possible typographic
errors in the taxonomic labels provided.

@@

From the list of namespaces in `prefixes`, we can follow links a handful
of established vocabularies and find terms that are not included in
`add_basic_meta()`.  For instance, by resolving the location identified
by the `prism` prefix,

@@

We can add this annotation to our existing `birds` NeXML file using the
`add_meta()` function.  Because we do not specify a level, it is added to
the root node, referring to the NeXML file as a whole.

@@

The `get_trees()` function can be used to return an `ape::phylo` tree as
usual.  `RNeXML` automatically detects the `simmap` reconstruction data
and returns includes this in a `maps` element of the `ape::phylo` object,
for use with other `phytools` functions.
